---
title: "Data Science Salaries Analysis Report"
author: "Piotr Sikora"
output: html_document
---

```{r libraries, include=FALSE}
library(tidyverse)
library(scales)
library(ggridges)
library(patchwork)

# Set theme for consistent visualizations
theme_set(theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    axis.title = element_text(size = 11),
    legend.position = "top"
  ))
```

```{r load-data, include=FALSE}
# Load the data
salaries_data <- read_csv("ds_salaries.csv")

# Data cleaning and preparation
salaries_data <- salaries_data %>%
  mutate(
    work_year = factor(work_year),
    experience_level = factor(
      experience_level,
      levels = c("EN", "MI", "SE", "EX"),
      labels = c("Entry", "Mid", "Senior", "Executive")
    ),
    employment_type = factor(
      employment_type,
      levels = c("PT", "FT", "CT", "FL"),
      labels = c("Part-Time", "Full-Time", "Contract", "Freelance")
    ),
    company_size = factor(
      company_size,
      levels = c("S", "M", "L"),
      labels = c("Small", "Medium", "Large")
    ),
    remote_ratio = factor(
      remote_ratio,
      levels = c(0, 50, 100),
      labels = c("On-site", "Hybrid", "Remote")
    ),
    # Categorize salary ranges for better visualization
    salary_category = cut(
      salary_in_usd,
      breaks = c(0, 50000, 100000, 150000, 200000, Inf),
      labels = c("<$50k", "$50-100k", "$100-150k", "$150-200k", "$200k+"),
      include.lowest = TRUE
    )
  )
```

# Introduction

This report examines compensation and employment trends in the data science field using a comprehensive salary dataset. Understanding salary structures, geographic variations, and the relationship between experience and compensation is crucial for both employers setting competitive offers and professionals navigating career decisions.

**Objective:** To analyze salary distributions across experience levels, company sizes, and geographic regions; examine employment patterns including remote work arrangements; and identify key factors influencing compensation in data science roles.

**Dataset Overview:** The dataset contains `r nrow(salaries_data)` employment records spanning multiple years with information on work year, experience level, employment type, job title, salary (in multiple currencies and standardized USD), employee residence, remote work ratio, company location, and company size.

# Salary Distributions

## Overall Salary Distribution

```{r distributions-salary-histogram, echo=FALSE, fig.height=7}
salary_stats <- salaries_data %>%
  summarise(
    mean_salary = mean(salary_in_usd),
    median_salary = median(salary_in_usd),
    sd_salary = sd(salary_in_usd)
  )

p1 <- ggplot(salaries_data, aes(x = salary_in_usd)) +
  geom_histogram(aes(y = after_stat(density)), 
                 bins = 50, fill = "#2E86AB", alpha = 0.7, 
                 color = "white", linewidth = 0.2) +
  geom_density(color = "#A23B72", linewidth = 1.2) +
  geom_vline(xintercept = salary_stats$median_salary, 
             linetype = "dashed", color = "darkred", linewidth = 1) +
  annotate("text", x = salary_stats$median_salary * 1.15, 
           y = Inf, 
           label = paste("Median:", dollar(salary_stats$median_salary, accuracy = 1)), 
           vjust = 1.5, hjust = 0, color = "darkred", fontface = "bold") +
  scale_x_continuous(labels = dollar_format(scale = 0.001, suffix = "k")) +
  labs(
    title = "Distribution of Data Science Salaries",
    subtitle = paste0("Mean: ", dollar(salary_stats$mean_salary, accuracy = 1), 
                     " | Median: ", dollar(salary_stats$median_salary, accuracy = 1),
                     " | SD: ", dollar(salary_stats$sd_salary, accuracy = 1)),
    x = "Annual Salary (USD)",
    y = "Density"
  )

p2 <- ggplot(salaries_data, aes(x = salary_in_usd, y = "")) +
  geom_boxplot(fill = "#2E86AB", alpha = 0.7, outlier.color = "#A23B72", 
               outlier.alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "darkred") +
  scale_x_continuous(labels = dollar_format(scale = 0.001, suffix = "k")) +
  labs(
    title = "Salary Distribution with Outliers",
    subtitle = "Diamond indicates mean; outliers shown in red",
    x = "Annual Salary (USD)"
  ) +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank()) 
p1 / p2
```

The salary distribution shows right-skewed characteristics with a median of `r dollar(salary_stats$median_salary, accuracy = 1)` and mean of `r dollar(salary_stats$mean_salary, accuracy = 1)`. The positive skew indicates a concentration of positions in the lower-to-middle salary ranges with a long tail of high-compensation roles, typical of data science markets with significant variation in role seniority and specialization.

## Salary Distribution by Experience Level

```{r distributions-experience, echo=FALSE, fig.height=7}
ggplot(salaries_data, aes(x = salary_in_usd, y = experience_level, 
                          fill = experience_level)) +
  geom_density_ridges(alpha = 0.8, scale=0.9, bandwidth=15000) +
  scale_fill_manual(values = c("Entry" = "#90E39A", "Mid" = "#5E9ECE",
                                "Senior" = "#F7B32B", "Executive" = "#E07A5F")) +
  scale_x_continuous(labels = dollar_format(scale = 0.001, suffix = "k")) +
  labs(
    title = "Salary Distributions Across Experience Levels",
    subtitle = "Clear progression in compensation with career advancement",
    x = "Annual Salary (USD)",
    y = "Experience Level",
    fill = "Experience"
  ) +
  theme(legend.position = "none")
```

Ridge plots reveal distinct salary distributions for each experience tier. Entry-level positions cluster around 60-80k dollars, mid-level around 80-120k dollars, senior positions show broader distribution centered near 140-160k dollars, and executive roles display the widest range with peaks above 180k dollars. The increasing spread at higher levels reflects greater variation in responsibilities and organizational contexts.

## Comparative Analysis by Experience and Company Size

```{r distributions-violin, echo=FALSE, fig.height=7}
sample_sizes <- salaries_data %>%
  count(experience_level, company_size) %>%
  mutate(label = paste0("n=", n))

ggplot(salaries_data, aes(x = company_size, y = salary_in_usd, 
                          fill = company_size)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.15, fill = "white", alpha = 0.8, 
               outlier.alpha = 0.3, outlier.size = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2.5, 
               fill = "red", color = "darkred") +
  scale_fill_manual(values = c("Small" = "#90E39A", "Medium" = "#5E9ECE", 
                                "Large" = "#A23B72")) +
  scale_y_continuous(labels = dollar_format(scale = 0.001, suffix = "k")) +
  facet_wrap(~experience_level, nrow = 1) +
  labs(
    title = "Salary Distribution by Company Size and Experience Level",
    subtitle = "Violin plots with embedded boxplots; diamonds indicate means",
    x = "Company Size",
    y = "Annual Salary (USD)"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "gray95", color = NA),
    strip.text = element_text(face = "bold")
  )
```

The interaction between company size and experience level reveals nuanced patterns. At entry and mid levels, company size shows modest impact on salary distributions. However, at senior and executive levels, medium and large companies show higher median salaries and greater spread, likely reflecting more complex organizational structures and higher-value projects at scale. Suprisingly, highest salaries for the executive level employees can be found in small sized companies. This might stem from overall smaller executive branches in such companies, which means the "big bucks" are split between less employees and also from the fact that bigger profits in smaller companies transfer to higher earnings for the executive suite more easily than in medium and large companies.

# Employment Patterns

## Distribution of Employment Types

```{r proportions-employment, echo=FALSE, fig.height=5}
employment_summary <- salaries_data %>%
  count(employment_type) %>%
  mutate(
    prop = n / sum(n),
    percentage = prop * 100
  ) %>%
  arrange(desc(n))

ggplot(employment_summary, aes(x = fct_reorder(employment_type, n), y = n)) +
  geom_col(fill = "#2E86AB", alpha = 0.85) +
  geom_text(aes(label = paste0(comma(n), "\n(", round(percentage, 1), "%)")),
            hjust = -0.1, size = 4) +
  coord_flip() +
  scale_y_continuous(
    labels = comma_format(),
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title = "Distribution of Employment Types in Data Science",
    subtitle = "Full-time positions dominate the employment landscape",
    x = NULL,
    y = "Number of Positions"
  )
```

Full-time employment overwhelmingly dominates the data science field, accounting for over 99% of positions in the dataset. Contract, freelance, and part-time arrangements remain rare, suggesting that organizations prefer traditional employment relationships for data science roles, likely due to the need for sustained engagement with complex projects and organizational knowledge.

## Remote Work Arrangements

```{r proportions-remote-stacked, echo=FALSE, fig.height=6}
remote_exp <- salaries_data %>%
  count(experience_level, remote_ratio) %>%
  group_by(experience_level) %>%
  mutate(
    prop = n / sum(n),
    percentage = prop * 100
  ) %>%
  ungroup()

ggplot(remote_exp, aes(x = experience_level, y = prop, fill = remote_ratio)) +
  geom_col(position = "fill", alpha = 0.85) +
  geom_text(
    aes(label = if_else(percentage > 8, 
                        paste0(round(percentage, 0), "%"), "")),
    position = position_fill(vjust = 0.5),
    size = 4,
    color = "white",
    fontface = "bold"
  ) +
  scale_y_continuous(
    labels = percent_format(),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c("On-site" = "#E07A5F", "Hybrid" = "#F7B32B", "Remote" = "#5E9ECE")
  ) +
  labs(
    title = "Remote Work Distribution by Experience Level",
    subtitle = "Remote work prevalence increases with seniority",
    x = "Experience Level",
    y = "Proportion",
    fill = "Work\nArrangement"
  )
```

Remote work adoption shows interesting variation across experience levels. Entry-level positions show the highest proportion of remote and hybrid work, likely reflecting the recent change in how entry level workers prioritize work from home. Fully on-site work becomes increasingly common at mid and executive levels, with fully on-site arrangements most prevalent among senior positions. Overall, the split between on-site or hybrid and remote workers at all levels is roughly 45/55.

# Temporal and Geographic Trends

## Salary Trends Over Time

```{r trends-time, echo=FALSE, fig.height=6}
yearly_stats <- salaries_data %>%
  group_by(work_year, experience_level) %>%
  summarise(
    mean_salary = mean(salary_in_usd),
    median_salary = median(salary_in_usd),
    se = sd(salary_in_usd) / sqrt(n()),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 10)  # Only show categories with sufficient data

ggplot(yearly_stats, aes(x = work_year, y = mean_salary, 
                         color = experience_level, group = experience_level)) +
  geom_line(linewidth = 1.2, alpha = 0.8) +
  geom_point(size = 3, alpha = 0.9) +
  geom_errorbar(aes(ymin = mean_salary - 1.96 * se, 
                    ymax = mean_salary + 1.96 * se),
                width = 0.2, alpha = 0.6, linewidth = 0.8) +
  scale_color_manual(values = c("Entry" = "#90E39A", "Mid" = "#5E9ECE",
                                 "Senior" = "#F7B32B", "Executive" = "#E07A5F")) +
  scale_y_continuous(labels = dollar_format(scale = 0.001, suffix = "k")) +
  labs(
    title = "Average Salary Trends by Experience Level Over Time",
    subtitle = "Error bars represent 95% confidence intervals",
    x = "Work Year",
    y = "Mean Annual Salary (USD)",
    color = "Experience\nLevel"
  )
```

Temporal analysis reveals evolving compensation trends across the years covered in the dataset. The inclusion of error bars representing 95% confidence intervals indicates the statistical uncertainty around these estimates. All experience levels show year-to-year variation, with executive and senior roles maintaining higher absolute compensation. Variation of compensation across experience levels dialed down since 2022, likely due to the recent tech boom slowing down after COVID-19 pandemic ended.

## Top Job Titles by Volume

```{r amounts-titles, echo=FALSE, fig.height=8, results='hide', fig.keep='all',warning=FALSE}
# Get top 15 job titles
top_titles <- salaries_data %>%
  count(job_title, sort = TRUE) %>%
  slice_head(n = 15)

title_data <- salaries_data %>%
  filter(job_title %in% top_titles$job_title) %>%
  group_by(job_title) %>%
  summarise(
    count = n(),
    mean_salary = mean(salary_in_usd),
    .groups = "drop"
  )

ggplot(title_data, aes(x = fct_reorder(job_title, count), y = count)) +
  geom_segment(aes(xend = fct_reorder(job_title, count), y = 0, yend = count), 
               color = "#2E86AB", size = 1) +
  geom_point(color = "#2E86AB", alpha = 1, size=5) +
  geom_text(aes(label = comma(count)), hjust = -1, size = 3) +
  coord_flip() +
  scale_y_continuous(
    labels = comma_format(),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Most Common Data Science Job Titles",
    subtitle = "Top 15 positions by total count in dataset",
    x = NULL,
    y = "Number of Positions"
  )
```

Job title analysis reveals the diversity of roles within data science. Data Scientist, Data Engineer, and Data Analyst comprise the most common positions, collectively representing the core of data science teams. More specialized roles like Machine Learning Engineer and Analytics Engineer also show substantial representation, reflecting the field's maturation and specialization.

## Geographic Distribution of Salaries

```{r distributions-geographic, echo=FALSE, fig.height=8}
# Focus on countries with sufficient data
top_countries <- salaries_data %>%
  count(company_location, sort = TRUE) %>%
  filter(n >= 30) %>%
  slice_head(n = 10) %>%
  pull(company_location)

country_data <- salaries_data %>%
  filter(company_location %in% top_countries)

ggplot(country_data, aes(x = fct_reorder(company_location, salary_in_usd, .fun = median), 
                         y = salary_in_usd)) +
  geom_boxplot(fill = "#2E86AB", alpha = 0.7, outlier.color = "#A23B72", 
               outlier.alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", size = 3, color= "darkred") +
  coord_flip() +
  scale_y_continuous(labels = dollar_format(scale = 0.001, suffix = "k")) +
  labs(
    title = "Salary Distribution by Company Location",
    subtitle = "Top 10 countries by number of positions; diamonds indicate means",
    x = "Company Location (Country Code)",
    y = "Annual Salary (USD)"
  )
```

Geographic analysis of compensation shows substantial variation across countries. US and Canada-based positions typically show higher median salaries and wider interquartile ranges, reflecting the mature tech market and high cost of living in major tech hubs. Other countries display more compressed salary ranges with lower medians, though some show competitive compensation at upper quartiles.

# Compensation Analysis

## Salary Distribution by Company Size

```{r amounts-company-size, echo=FALSE, fig.height=6}
company_summary <- salaries_data %>%
  group_by(company_size) %>%
  summarise(
    mean_salary = mean(salary_in_usd),
    median_salary = median(salary_in_usd),
    count = n(),
    .groups = "drop"
  )

ggplot(company_summary, aes(x = company_size, y = mean_salary, fill = company_size)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = paste0(dollar(mean_salary, accuracy = 1), "\n(n=", 
                                comma(count), ")")),
            vjust = -0.5, size = 4) +
  scale_y_continuous(
    labels = dollar_format(scale = 0.001, suffix = "k"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_fill_manual(values = c("Small" = "#90E39A", "Medium" = "#5E9ECE", 
                                "Large" = "#A23B72")) +
  labs(
    title = "Average Salary by Company Size",
    subtitle = "Mean compensation with sample sizes",
    x = "Company Size",
    y = "Mean Annual Salary (USD)"
  ) +
  theme(legend.position = "none")
```

Company size shows suprising association with average compensation. Medium-sized companies offer the highest mean salaries, while large companies show comparable or slightly lower compensation. As expected, small-sized companies show the lowest mean compensation. The differences are relatively small in magnitude, suggesting that factors like location, industry, and specific role may matter more than organizational size alone.

## Remote Work Impact on Compensation

```{r associations-remote-salary, echo=FALSE, fig.height=6}
remote_salary <- salaries_data %>%
  group_by(remote_ratio, experience_level) %>%
  summarise(
    mean_salary = mean(salary_in_usd),
    median_salary = median(salary_in_usd),
    se = sd(salary_in_usd) / sqrt(n()),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 15)

ggplot(remote_salary, aes(x = experience_level, y = mean_salary, 
                          fill = remote_ratio, group = remote_ratio)) +
  geom_col(position = "dodge", alpha = 0.85) +
  geom_errorbar(
    aes(ymin = mean_salary - 1.96 * se, ymax = mean_salary + 1.96 * se),
    position = position_dodge(width = 0.9),
    width = 0.25,
    linewidth = 0.8,
    alpha = 0.7
  ) +
  scale_y_continuous(
    labels = dollar_format(scale = 0.001, suffix = "k"),
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_fill_manual(
    values = c("On-site" = "#E07A5F", "Hybrid" = "#F7B32B", "Remote" = "#5E9ECE")
  ) +
  labs(
    title = "Average Salary by Experience Level and Remote Work Status",
    subtitle = "Error bars show 95% confidence intervals",
    x = "Experience Level",
    y = "Mean Annual Salary (USD)",
    fill = "Work\nArrangement"
  )
```

The relationship between remote work arrangements and compensation shows nuanced patterns across experience levels. At entry and mid levels, on-site positions trend slightly higher in mean compensation, possibly reflecting geographic concentration in high-cost markets. At senior and executive levels, remote positions show competitive or higher compensation, suggesting these roles command premium pay regardless of location due to scarce specialized talent.

## Salary Categories by Experience

```{r proportions-salary-categories, echo=FALSE, fig.height=6}
salary_cat_exp <- salaries_data %>%
  count(experience_level, salary_category) %>%
  group_by(experience_level) %>%
  mutate(
    prop = n / sum(n),
    percentage = prop * 100,
    salary_category = fct_rev(salary_category)
  ) %>%
  ungroup()

ggplot(salary_cat_exp, aes(x = experience_level, y = prop, 
                            fill = salary_category)) +
  geom_col(position = "fill", alpha = 0.85) +
  geom_text(
    aes(label = if_else(percentage > 5, 
                        paste0(round(percentage, 0), "%"), "")),
    position = position_fill(vjust = 0.5),
    size = 3.5,
    color = "white",
    fontface = "bold"
  ) +
  scale_y_continuous(
    labels = percent_format(),
    expand = c(0, 0)
  ) +
  scale_fill_viridis_d(option = "plasma", direction = 1) +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(
    title = "Salary Range Distribution by Experience Level",
    subtitle = "Clear progression into higher salary brackets with seniority",
    x = "Experience Level",
    y = "Proportion",
    fill = "Salary Range"
  )
```

Salary category distributions demonstrate clear career progression patterns. Entry-level positions concentrate heavily in the <50k and 50-100k ranges, with some representation above 150k. Senior roles show more balanced distribution across middle categories, while executive positions predominantly occupy the 150k+ brackets, with significant representation above 200k. This progression validates expected career earnings trajectories.

# Summary and Conclusions

This comprehensive analysis of data science employment and compensation reveals several critical insights:

**Salary Landscape:**

- Overall median salary of `r dollar(salary_stats$median_salary, accuracy = 1)` with right-skewed distribution
- Clear salary progression: Entry (~70k) → Mid (~110k) → Senior (~150k) → Executive (~200k)
- Company size shows modest impact; geographic location and experience level are stronger predictors

**Employment Structure:**

- Full-time employment dominates (>99%), with minimal contract or freelance arrangements
- Remote work increases with seniority, reaching 40-50% at senior/executive levels
- All level positions are 45/55 when it comes to on-site presence, with 55% being remote/hybrid.

**Market Dynamics:**

- Substantial geographic variation in compensation, with US/CA positions commanding premium salaries
- Temporal trends show year-over-year variation with experience level differentiation maintained
- Job title diversity reflects field specialization, with Data Scientist, Engineer, and Analyst as core roles

**Key Relationships:**

- Experience level is the strongest determinant of compensation
- Remote work arrangements show neutral-to-positive compensation impact at senior and executive levels
- Medium-sized companies offer competitive compensation compared to large enterprises

**Professional Implications:**

For professionals: Focus on advancing experience level as the primary salary lever; geographic and company size considerations are secondary. Remote opportunities become more available and competitive at senior levels.

For employers: Compensation must align with experience tiers to remain competitive. Remote work flexibility is increasingly expected at all levels without necessarily commanding premium pay.

The data science employment market shows maturation with clear career ladders, though substantial geographic and organizational variation persists. The field's evolution toward remote work and specialization suggests continued dynamic adaptation in coming years.

---

**Methodological Notes:** Analysis includes 95% confidence intervals on temporal and grouped estimates to represent statistical uncertainty. Sample size filters applied to geographic analyses to ensure reliable estimates.